### 2016-11-23 Manuscript/SW_early_life_telomeres.Rmd
Home eds

#####Statistical analyses
We performed all statistical analyses using R. We first tested the relationship between RTL and age using linear mixed models (LMMs) with individual ID as a random effect, carried out in the lme4 package [@Bates2014]. To test whether the decrease in telomere length with age is greatest in early life, we compared LMMs with age (in months) and log-transformed age, using a likelihood ratio test. We also tested whether telomores shorten with age in individuals with samples from multiple timepoints, using a LMM of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009]. To test how telomere length, and its relationship with age, varied among cohorts, we compared the above LMMs with models that included birth year and catch year as additional random effects. We also compared a model in which only the intercept (i.e. RTL) was allowed to vary among among cohorts or years, to one in which the slope of the relationship between RTL and age was also allowed to vary.

Because we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error. Using all birds (adults and juveniles) with at least two telomere measurement, we calculated $\Delta$RTL for each pair of samples (for example, for a bird with three longitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample). We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (i.e. completely seperate reactions run on separate plates: N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples). We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at the qPCR was run rather than by age. We first compared the variance in $\Delta$RTL within and among individuals using a Levene's test. Then, to separately test whether the extent of telomere shortening and lenghtening within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which shortening ($\Delta$RTL < 0) and lengthening $\Delta$RTL ($\Delta$RTL > 0) occurred, and tested whether $\Delta$RTL among samples was significantly different between within-sample and across-sample measurements, using Wilcoxon tests.

>>>>

#####Statistical analyses
We performed all statistical analyses using R. We first explored the relationship between RTL and age among cohorts using linear mixed models (LMMs) carried out in the lme4 package [@Bates2014]. Following a similar approach to Fairlie *et al.* [-@Fairlie2016], we compared a selection of models fitting different relationships between RTL and age. We created models where age was fitted as a linear covariate, where age was log-transformed, and where age was fitted as a factor. We also created models including birth year (cohort) as a factor, and ones which included interactions between cohort and each measure of age. All fitted models are included in Table 1. As random effects we included individual ID, catch year and qPCR plate ID. Models were compared using AIC.

Using the longitudinal data, we tested whether telomores shorten with age in individuals, using a LMM of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009]. Because we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error. Using all birds (adults and juveniles) with at least two telomere measurement, we calculated $\Delta$RTL for each pair of samples (for example, for a bird with three longitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample). We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (i.e. completely seperate reactions run on separate plates: N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples). We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at the qPCR was run rather than by age. We first compared the variance in $\Delta$RTL within and among individuals using a Levene's test. Then, to separately test whether the extent of telomere shortening and lenghtening within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which shortening ($\Delta$RTL < 0) and lengthening $\Delta$RTL ($\Delta$RTL > 0) occurred, and tested whether $\Delta$RTL among samples was significantly different between within-sample and across-sample measurements, using Wilcoxon tests.

### 2016-11-23 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home eds

**Table 1**
```{r Table 1}

AIC(RTL_LogAge)
AIC(RTL_LogAge_LY)

@@

CIs <- CIwald(RTL_Full)
t1 <- data.frame(Estimate = fixef(RTL_Full), LCI = CIs[7:nrow(CIs),1], UCI = CIs[7:nrow(CIs),2])

nams <- row.names(t1)
nams[nams == 'LogAge'] <- 'Age'
nams[nams == 'Tarsus'] <- 'Tarsus length'
nams[nams == 'Helper'] <- 'Number helpers'
nams[nams == 'GroupSize'] <- 'Group size'
nams[nams == 'SexMales'] <- 'Sex (males)'
nams[nams == 'BodyMass'] <- 'Body mass'
nams[nams == 'Insect'] <- 'Annual food availability'
nams[nams == 'TQ'] <- 'Territory quality'

row.names(t1) <- nams

kable(t1)

**Table 2**

```{r Table 2}

#Run the model
CIs <- CIwald(DeltaRTL_Full)
t1 <- data.frame(Estimate = fixef(DeltaRTL_Full), LCI = CIs[4:nrow(CIs),1], UCI = CIs[4:nrow(CIs),2])

nams <- row.names(t1)
nams[nams == 'LogAge'] <- 'Age'
nams[nams == 'Tarsus'] <- 'Tarsus length'
nams[nams == 'Helper'] <- 'Number helpers'
nams[nams == 'GroupSize'] <- 'Group size'
nams[nams == 'SexMales'] <- 'Sex (males)'
nams[nams == 'BodyMass'] <- 'Body mass'
nams[nams == 'Insect'] <- 'Annual food availability'
nams[nams == 'TQ'] <- 'Territory quality'

row.names(t1) <- nams

kable(t1)

>>>>

**Table 1**
```{r Table 1}
RTL_Age_Dredge <- dredge(RTL_Age,
                         subset = !(LogAge & FAge) & 
                           !(LogAge & Agemonths) &
                           !(LogAge & Age2) &
                           !(Agemonths & FAge) & 
                           !(Agemonths & Age2) &
                           !(FAge & Age2))

t1 <- as.matrix(RTL_Age_Dredge)

#Sort out row/col names
nams <- colnames(t1)

nams[nams == 'Agemonths'] <- 'Age (linear)'
nams[nams == 'LogAge'] <- 'Age (log)'
nams[nams == 'FAge'] <- 'Age (factor)'
nams[nams == 'Cohort:FAge'] <- 'Age (factor)*cohort'
nams[nams == 'Agemonths:Cohort'] <- 'Age (continuous)*cohort'
nams[nams == 'Cohort:LogAge'] <- 'Age (log)*cohort'
nams[nams == 'Age2:Cohort'] <- 'Age (quadratic)*cohort'
nams[nams == 'Age2'] <- 'Age (quadratic)'

colnames(t1) <- nams

tableS1A <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
colnames(tableS1A) <- c('Model','df','AICc', 'Delta AICc','Weight')
rownames(tableS1A) <- NULL
tableS1A$Model[tableS1A$Model == ''] <- 'Null model'
tableS1A[,3:5] <- lapply(tableS1A[,3:5],as.numeric)

RTL_DeltaAge_Dredge <- dredge(RTL_DeltaAge,
                         subset = !(DeltaLogAge & DeltaAge) &
                           !(DeltaLogAge & DeltaAge2) &
                           !(DeltaAge & DeltaAge2))

t1 <- as.matrix(RTL_DeltaAge_Dredge)

#Sort out row/col names
nams <- colnames(t1)

nams[nams == 'DeltaAge'] <- 'Delta age (linear)'
nams[nams == 'DeltaLogAge'] <- 'Delta age (log)'
nams[nams == 'Cohort:DeltaAge'] <- 'Delta age (continuous)*cohort'
nams[nams == 'Cohort:DeltaLogAge'] <- 'Delta age (log)*cohort'
nams[nams == 'Cohort:DeltaAge2'] <- 'Delta age (quadratic)*cohort'
nams[nams == 'DeltaAge2'] <- 'Delta age (quadratic)'

colnames(t1) <- nams

tableS1B <- data.frame(Model = apply(t1,1,getterms,c(2:8)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
colnames(tableS1B) <- c('Model','df','AICc', 'Delta AICc','Weight')
rownames(tableS1B) <- NULL
tableS1B$Model[tableS1B$Model == ''] <- 'Null model'
tableS1B[,3:5] <- lapply(tableS1B[,3:5],as.numeric)

tableS1 <- rbind(c('A',rep(' ',4)),tableS1A,c('--',rep(' ',4)),c('B',rep(' ',4)),tableS1B)

kable(tableS1,digits = 3)
```

@@

CIs <- CIwald(RTL_Full)
t1a <- data.frame(Variable = row.names(CIs)[7:nrow(CIs)],
                  Estimate = fixef(RTL_Full),
                  LCI = CIs[7:nrow(CIs),1],
                  UCI = CIs[7:nrow(CIs),2],
                  row.names = NULL,
                  stringsAsFactors = F)

nams <- t1a$Variable
nams[nams == 'LogAge'] <- 'Age'
nams[nams == 'Tarsus'] <- 'Tarsus length'
nams[nams == 'Helper'] <- 'Number helpers'
nams[nams == 'GroupSize'] <- 'Group size'
nams[nams == 'SexMales'] <- 'Sex (males)'
nams[nams == 'BodyMass'] <- 'Body mass'
nams[nams == 'Insect'] <- 'Annual food availability'
nams[nams == 'TQ'] <- 'Territory quality'

t1a$Variable <- nams

#Run the model
CIs <- CIwald(DeltaRTL_Full)
t1b <- data.frame(Variable = row.names(CIs)[4:nrow(CIs)],
                  Estimate = fixef(DeltaRTL_Full),
                  LCI = CIs[4:nrow(CIs),1],
                  UCI = CIs[4:nrow(CIs),2],
                  row.names = NULL,
                  stringsAsFactors = F)

nams <- t1b$Variable
nams[nams == 'LogAge'] <- 'Age'
nams[nams == 'Tarsus'] <- 'Tarsus length'
nams[nams == 'Helper'] <- 'Number helpers'
nams[nams == 'GroupSize'] <- 'Group size'
nams[nams == 'SexMales'] <- 'Sex (males)'
nams[nams == 'BodyMass'] <- 'Body mass'
nams[nams == 'Insect'] <- 'Annual food availability'
nams[nams == 'TQ'] <- 'Territory quality'

t1b$Variable <- nams

tableS2 <- rbind(c('A',rep(' ',4)),t1a,c('--',rep(' ',4)),c('B',rep(' ',4)),t1b)

kable(tableS2)
